Supplementary materials for ``Inattentive responding can induce spurious associations between task behavior and symptom measures''

### 2023-04-27 manuscripts/main.tex
reupload

% make abstract
\abstract{
\noindent Although online samples have many advantages for psychiatric research, some potential pitfalls of this approach are not widely understood. Here, we detail circumstances in which spurious correlations may arise between task behavior and symptom scores. The problem arises because many psychiatric symptom surveys have asymmetric score distributions in the general population, meaning that careless responders on these surveys will show apparently elevated symptom levels. If these participants are similarly careless in their task performance, this may result in a spurious association between symptom scores and task behavior. We demonstrate this pattern of results in two samples of participants (total N = 779) recruited online who performed one of two common cognitive tasks. False-positive rates for these spurious correlations increase with sample size, contrary to common assumptions. Excluding participants flagged for careless responding on surveys abolished these spurious correlations, but exclusion based on task performance alone was less effective.
}

@@

The present study highlights the need for future work on the prevalence of C/IE responding in online samples and its interactions with task-symptom correlations. Many open questions remain, including under what conditions task- and symptom-screening measures might better correspond, what screening measures are most effective and when, and under what conditions spurious correlations are more likely to arise. For example, in the present study, we found that screening on task behavior alone was insufficient to prevent putatively spurious correlations for one task (reversal learning) but not for another (two-step task). This discrepancy may reflect differences in the tasks (e.g., the two-step task may be more challenging and thus more sensitive to C/IE responding) or differences in the screening measures (e.g., choice accuracy across 90 trials may be a noisier measure than win-stay lose-shift choice behavior across 200 trials). One especially pressing question is how sample size affects the likelihood of obtaining spurious correlations. The results of a bootstrapping analysis in our data suggest that false positive rates are likely to increase with sample size. As computational psychiatry studies move towards larger samples to characterize heterogeneity in symptoms (and to increase statistical power), it will be important to understand how sample size may exaggerate the effects of systematic error. It will also be important to understand how this is moderated by overall C/IE responding rates, which we observed to vary across platforms and time, and which will presumably continue to evolve with changing labor platform and researcher screening practices.

We conclude with a list of concrete recommendations for future online studies involving correlations between task behavior and self-report instruments. We note that these recommendations are not limited to computational psychiatry studies, but are applicable to any online individual-differences cognitive science research involving similar methods (e.g., behavioral economics, psycholinguistics). Moving forward, we would strongly recommend that experimenters employ some form of self-report screening method, preferably one recommended by the best-practices literature (e.g., \cite{huang2015insufficient, curran2016methods, thomas2017validity, desimone2018dirty, ophir2020turker}). Our literature review found that, to date, the majority of online studies assessing behavior-symptom correlations have not used self-report screening, and our results demonstrate that stand-alone task-behavior screening is often insufficent to prevent spurious symptom-behavior correlations induced by C/IE responding. For these reasons, we encourage experimenters to use a variety of data-quality checks for online studies and to be transparent in their reporting of how screening was conducted, how many participants are flagged under each measure, and what thresholds are used for rejection.

When collecting self-report questionnaire data, we encourage experimenters to use screening methods sensitive to multiple distinct patterns of C/IE responding (e.g., random responding, straight-lining, side bias) and, if possible, to log all page interactions (e.g., mouse clicks, keyboard presses). We specifically recommend experimenters follow best-practice guidelines and use infrequency-item attention checks rather than instructed-item checks, as multiple studies have now shown that online participants are habituated to and circumvent the latter (e.g., \cite{barends2019noncompliant, thomas2017validity, hauser2016attentive}; Appendix B). Participants flagged by suspicious responses on attention-check items should either be excluded from further analysis, or assessed using sensitivity analyses to ensure that observed full-sample correlations are robust to their exclusion.

We found that spurious correlations predominantly affected self-report instruments for which the expected distributions of symptom scores were asymmetric (either positively or negatively skewed). As such, symmetrically-distributed measures of a given construct should be preferred to asymmetrically-distributed measures (though this will often be infeasible given that the prevalence of many psychiatric symptoms in the general population is typically small). Scales with reverse-coded items can be used to quantify the consistency of participants' responses between reverse-coded and non-reverse-coded measures of the same latent construct. With some care, this may be used to identify C/IE responding even for measures that do not include attention-check items \cite{emons2009detection}. Similarly, it may be beneficial to include multiple self-report surveys of the same construct to measure consistency across scales.

In our experience, we have found it instructive to review discussions on public forums for participants of online labor markets (e.g., Reddit, TurkerNation). Doing so helps an experimenter identify what screening methods would-be participants are already aware of and prepared to answer correctly. (Several examples of workers discussing common attention checks can be found at the Github repository for this project.)

@@

409 total participants were recruited to participate in an online behavioral experiment in late June - early July, 2020. Specifically, 208 participants were recruited from Amazon Mechanical Turk (MTurk) and 201 participants were recruited from Prolific. This study was approved by the Institutional Review Board of Princeton University (\#5291), and all participants provided informed consent. Total study duration was approximately 10 minutes per participant. Participants received monetary compensation for their time (rate USD \$12/hr), plus an incentive-compatible bonus up to \$0.25 based on task performance.

@@

The data from multiple participants who completed the experiment were excluded prior to analysis. Three participants (all MTurk) were excluded due to missing data. In addition, we excluded 20 participants who disclosed that they had also completed the experiment on the other platform. This left a final sample of N=386 participants (MTurk: N=186, Prolific: N=200) for analysis. The demographics of the sample split by labor market is provided in Table S1. Notably, the participants recruited from MTurk were older ($M = 7.7 \ \text{yrs}$, two-tailed, two-sample $t$-test: $t(384)=6.567$, $p<0.001$, $d=0.669$, 95\% CI = [5.4, 10.0]) and comprised of fewer women (two-tailed, two-sample proportions test: $z(384)=2.529$, $p=0.011$, $h=0.258$, 95\% CI = [0.030, 0.228]).

@@

The authors are grateful to Agnes Norbury, Alexandra Pike, and Oliver Robinson for helpful discussion. The research reported in this manuscript was supported in part by the National Institute of Mental Health (NIMH) under award number 5R01MH119511-02 (YN), and by the National Center for Advancing Translational Sciences (NCATS), a component of the National Institute of Health (NIH), under award number UL1TR003017 (YN). The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health. SZ was supported by an NSF Graduate Research Fellowship. DB was supported by an Early Career Fellowship from the Australian National Health and Medical Research Council (\#1165010). The funders had no role in study design, data collection and analysis, decision to publish or preparation of the manuscript.

@@

\begin{figure}[h]
    % \includegraphics[width=16cm]{../figures/main_01.png}
    \centering
    \caption{Simulated example of how spurious behavior-symptom correlations can arise when symptom endorsement is rare. \textit{Left:} When symptoms are moderately common in the general population, C/IE respondents (blue) are indistinguishable from attentive participants (red) in self-report measures (x-axis, marginal distribution shown on top). Despite the worse task performance of C/IE respondents (y-axis), no correlation arises between symptom scores and task performance (dots are participants drawn from the shown distributions, with 15\% C/IE participants; dashed line shows the (lack of) Spearman rank correlation.) \textit{Right:} When symptoms are rare in the general population, careless respondents appear symptomatic in self-report measures. As a result, self-report symptom scores show a significant Spearman rank correlation (two-sided) with task performance.}
    \label{fig:simulation}
\end{figure}

\clearpage
\begin{figure}[h]
    % \includegraphics[width=16cm]{../figures/main_02a.png}
    \centering
    \caption{Raincloud plots of total symptom scores in attentive ($N=301$; red) and C/IE ($N=85$; blue) participants. Each colored dot represents the symptom score for one participant. Black circles: average score within each group (error bars denote 95\% bootstrap confidence interval). Shaded plots: distribution of scores for each group of participants. The scales are ordered approximately according to their estimated skew (see Table \ref{tab:tab02}) from top-left (7-up) to bottom-right (PSWQ). The average level of symptom endorsement is most markedly different between groups in symptom measures with the lowest overall rates of endorsement.}
    \label{fig:fig01}
\end{figure}

\clearpage
\begin{figure}[h]
    % \includegraphics[width=16cm, trim=0 0 3cm 0]{../figures/main_02b.png}
    \centering
    \caption{Similarity of task and self-report data screening measures. Each tile corresponds to the Spearman rank correlation (left) and Dice similarity coefficient (right) between two screening measures across participants ($N=386$). Similarity indices are thresholded such that only the magnitude of statistically-significant associations (permutation test, $p<0.05$, two-sided, corrected for multiple comparisons) are shown. (Unthresholded values are presented in Tables S3--S5.) Cross-modality correlations between task (y-axis) and self-report screening measures (x-axis) are in the dashed rectangle. Acronyms: INF = infrequency item; ISD = inter-item standard deviation; REL = personal reliability; MAH = Mahalanobis distance; READ = reading time; VAR = choice variability; ACC = choice accuracy; WSLS = win-stay lose-shift rate; RT = suspicious response times.}
    \label{fig:fig02}
\end{figure}

\clearpage
\begin{figure}[h]
    % \includegraphics[width=17cm]{../figures/main_03a.png}
    \centering
    \caption{Absolute Spearman rank correlations between task behavior (y-axis) and symptom measures (x-axis) under different regimes of data screening and participant exclusions. (A) No Screening = no exclusions ($N=386$). (B) Accuracy Only = exclusions based on chance-level performance in the reversal-learning task ($N=352$). (C) Infrequency Only = exclusions based on invalid or improbable responses to infrequency items ($N=301$). (D) Both Types = exclusions based on the previous two measures ($N=283$). Only statistically significant correlations are shown ($p<0.05$, two-sided, not corrected for multiple comparisons; signed correlations are shown in Figure S1 and Tables S6--S9). Black Xs indicate significant correlations abolished under screening. Acronyms: Acc = choice accuracy; Pts = total points earned; WS = win-stay rate; LS = lose-shift rate; Pers = perseveration errors; $\beta$ = inverse temperature; $\eta_p$ = positive learning rate; $\eta_n$ = negative learning rate; $\kappa$ = learning rate asymmetry.}
    \label{fig:fig03}
\end{figure}

\clearpage
\begin{figure}[h]
    % \includegraphics[width=16cm]{../figures/main_03b.png}
    \centering
    \caption{False positive rates for spurious correlations \emph{increase} with sample size. \textit{Left:} Spearman rank correlations and 95\% bootstrap confidence intervals between learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. The thick dashed lines indicate the threshold for statistical significance for the Spearman correlation at the corresponding sample size. Markers are jittered along the x-axis for legibility. \textit{Right:} False positive rates for learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. False positive rate was calculated as the proportion of bootstrap samples in which the Spearman rank correlation between $\kappa$ and 7-down was statistically significant ($p<0.05$, two-sided). The horizontal dotted line denotes the expected false positive rate at $\alpha = 0.05$.}
    \label{fig:fig04}
\end{figure}

>>>>

% make abstract
\abstract{
\noindent Although online samples have many advantages for psychiatric research, some potential pitfalls of this approach are not widely understood. Here, we detail circumstances in which spurious correlations may arise between task behavior and symptom scores. The problem arises because many psychiatric symptom surveys have asymmetric score distributions in the general population, meaning that careless responders on these surveys will show apparently elevated symptom levels. If these participants are similarly careless in their task performance, this may result in a spurious association between symptom scores and task behavior. We demonstrate this pattern of results in two samples of participants recruited online (total N = 779) who performed one of two common cognitive tasks. False-positive rates for these spurious correlations increase with sample size, contrary to common assumptions. Excluding participants flagged for careless responding on surveys abolished the spurious correlations, but exclusion based on task performance alone was less effective.
}

@@

This study highlights the need for more research on the prevalence of C/IE responding in online samples and its interactions with task-symptom correlations. Many open questions remain, including under what conditions task- and symptom-screening measures might better correspond, what screening measures are most effective and when, and under what conditions spurious correlations are more likely to arise. For example, we found that screening on task behavior alone was insufficient to prevent putatively spurious correlations for one task (reversal learning) but was sufficient for another task (the two-step task). This discrepancy may reflect differences in the tasks (e.g., the two-step task may be more challenging and thus more sensitive to C/IE responding) or differences in the screening measures (e.g., choice accuracy across 90 trials may be a noisier measure than win-stay lose-shift choice behavior across 200 trials).

One especially pressing question is how sample size affects the likelihood of obtaining spurious correlations. The results of a bootstrapping analysis in our data suggest that false positive rates are likely to increase with sample size. As computational psychiatry studies move towards larger samples to characterize heterogeneity in symptoms (and to increase statistical power), it will be important to understand how sample size may exaggerate the effects of systematic error. It will also be important to understand how this is moderated by overall C/IE responding rates, which we observed to vary across platforms and time, and which will presumably continue to evolve with changing labor platform and researcher screening practices.

We conclude with a list of concrete recommendations for future online studies involving correlations between task behavior and self-report instruments. We note that these recommendations are not limited to computational psychiatry studies, but are applicable to any online individual-differences cognitive science research involving similar methods (e.g., behavioral economics, psycholinguistics).

Moving forward, we strongly recommend that experimenters employ some form of self-report screening method, preferably one recommended by the best-practices literature (e.g., \cite{huang2015insufficient, curran2016methods, thomas2017validity, desimone2018dirty, ophir2020turker}). Our literature review found that, to date, the majority of online studies assessing behavior-symptom correlations have not used self-report screening, and our results demonstrate that stand-alone task-behavior screening is not necessarily sufficent to prevent spurious symptom-behavior correlations induced by C/IE responding. We therefore encourage experimenters to use a variety of data-quality checks for online studies and to be transparent in their reporting of how screening was conducted, how many participants were flagged under each measure, and what thresholds were used for rejection.

When collecting self-report questionnaire data, we encourage experimenters to use screening methods sensitive to multiple distinct patterns of C/IE responding (e.g., random responding, straight-lining, side bias) and, if possible, to log all page interactions (e.g., mouse clicks, keyboard presses). We specifically recommend experimenters use infrequency-item attention checks rather than instructed-item checks, as multiple studies have now shown that online participants are habituated to and circumvent the latter (e.g., \cite{barends2019noncompliant, thomas2017validity, hauser2016attentive}; Appendix B). Participants flagged by suspicious responses on attention-check items should either be excluded from further analysis, or assessed using sensitivity analyses to ensure that observed full-sample correlations are robust to their exclusion.

We found that spurious correlations predominantly affected self-report instruments for which the expected distributions of symptom scores were asymmetric (either positively or negatively skewed). As such, all else equal, symmetrically-distributed measures of a given construct should be preferred to asymmetrically-distributed measures (though this will often be infeasible given that the prevalence of many psychiatric symptoms in the general population is typically small). Scales with reverse-coded items can be used to quantify the consistency of participants' responses between reverse-coded and non-reverse-coded measures of the same latent construct. With some care, this may be used to identify C/IE responding even for measures that do not include attention-check items \cite{emons2009detection}. Similarly, it may be beneficial to include multiple self-report surveys of the same construct to measure consistency across scales.

In our experience, we have found it instructive to review discussions on public forums for participants of online labor markets (e.g., at the time of writing, Reddit, TurkerNation). Doing so helps an experimenter identify what screening methods would-be participants are already aware of and prepared to answer correctly. (Several examples of workers discussing common attention checks can be found at the Github repository for this project.)

@@

409 total participants were recruited to participate in an online behavioral experiment in late June - early July, 2020. Specifically, 208 participants were recruited from Amazon Mechanical Turk (MTurk) and 201 participants were recruited from Prolific. This study was approved by the Institutional Review Board of Princeton University, and all participants provided informed consent. Total study duration was approximately 10 minutes per participant. Participants received monetary compensation for their time (rate USD \$12/hr), plus an incentive-compatible bonus up to \$0.25 based on task performance.

@@

Data from several participants were excluded prior to analysis. Three participants (all MTurk) were excluded due to missing data. In addition, we excluded 20 participants who disclosed that they had also completed the experiment on the other platform. This left a final sample of N=386 participants (MTurk: N=186, Prolific: N=200) for analysis. The demographics of the sample split by labor market is provided in Table S1. Notably, the participants recruited from MTurk were older ($M = 7.7 \ \text{yrs}$, two-tailed, two-sample $t$-test: $t(384)=6.567$, $p<0.001$, $d=0.669$, 95\% CI = [5.4, 10.0]) and comprised of fewer women (two-tailed, two-sample proportions test: $z(384)=2.529$, $p=0.011$, $h=0.258$, 95\% CI = [0.030, 0.228]).

@@

The authors are grateful to Agnes Norbury, Alexandra Pike, and Oliver Robinson for helpful discussion. Research reported in this manuscript was supported in part by the National Institute of Mental Health (R01MH119511; YN), and by the National Center for Advancing Translational Sciences (UL1TR003017; YN). The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health. SZ was supported by an NSF Graduate Research Fellowship. DB was supported by an Early Career Fellowship from the Australian National Health and Medical Research Council (\#1165010). The funders had no role in study design, data collection and analysis, decision to publish or preparation of the manuscript.

@@

\begin{figure}[h]
    \includegraphics[width=16cm]{../figures/main_01.png}
    \centering
    \caption{Simulated example of how spurious behavior-symptom correlations can arise when symptom endorsement is rare. \textit{Left:} When symptoms are moderately common in the general population, C/IE respondents (blue) are indistinguishable from attentive participants (red) in self-report measures (x-axis, marginal distribution shown on top). Despite the worse task performance of C/IE respondents (y-axis), no correlation arises between symptom scores and task performance (dots are participants drawn from the shown distributions, with 15\% C/IE participants; dashed line shows the (lack of) Spearman rank correlation.) \textit{Right:} When symptoms are rare in the general population, careless respondents appear symptomatic in self-report measures. As a result, self-report symptom scores show a significant Spearman rank correlation (two-sided) with task performance.}
    \label{fig:simulation}
\end{figure}

\clearpage
\begin{figure}[h]
    \includegraphics[width=16cm]{../figures/main_02a.png}
    \centering
    \caption{Raincloud plots of total symptom scores in attentive ($N=301$; red) and C/IE ($N=85$; blue) participants. Each colored dot represents the symptom score for one participant. Black circles: average score within each group (error bars denote 95\% bootstrap confidence interval). Shaded plots: distribution of scores for each group of participants. The scales are ordered approximately according to their estimated skew (see Table \ref{tab:tab02}) from top-left (7-up) to bottom-right (PSWQ). The average level of symptom endorsement is most markedly different between groups in symptom measures with the lowest overall rates of endorsement.}
    \label{fig:fig01}
\end{figure}

\clearpage
\begin{figure}[h]
    \includegraphics[width=16cm, trim=0 0 3cm 0]{../figures/main_02b.png}
    \centering
    \caption{Similarity of task and self-report data screening measures. Each tile corresponds to the Spearman rank correlation (left) and Dice similarity coefficient (right) between two screening measures across participants ($N=386$). Similarity indices are thresholded such that only the magnitude of statistically-significant associations (permutation test, $p<0.05$, two-sided, corrected for multiple comparisons) are shown. (Unthresholded values are presented in Tables S3--S5.) Cross-modality correlations between task (y-axis) and self-report screening measures (x-axis) are in the dashed rectangle. Acronyms: INF = infrequency item; ISD = inter-item standard deviation; REL = personal reliability; MAH = Mahalanobis distance; READ = reading time; VAR = choice variability; ACC = choice accuracy; WSLS = win-stay lose-shift rate; RT = suspicious response times.}
    \label{fig:fig02}
\end{figure}

\clearpage
\begin{figure}[h]
    \includegraphics[width=17cm]{../figures/main_03a.png}
    \centering
    \caption{Absolute Spearman rank correlations between task behavior (y-axis) and symptom measures (x-axis) under different regimes of data screening and participant exclusions. (A) No Screening = no exclusions ($N=386$). (B) Accuracy Only = exclusions based on chance-level performance in the reversal-learning task ($N=352$). (C) Infrequency Only = exclusions based on invalid or improbable responses to infrequency items ($N=301$). (D) Both Types = exclusions based on the previous two measures ($N=283$). Only statistically significant correlations are shown ($p<0.05$, two-sided, not corrected for multiple comparisons; signed correlations are shown in Figure S1 and Tables S6--S9). Black Xs indicate significant correlations abolished under screening. Acronyms: Acc = choice accuracy; Pts = total points earned; WS = win-stay rate; LS = lose-shift rate; Pers = perseveration errors; $\beta$ = inverse temperature; $\eta_p$ = positive learning rate; $\eta_n$ = negative learning rate; $\kappa$ = learning rate asymmetry.}
    \label{fig:fig03}
\end{figure}

\clearpage
\begin{figure}[h]
    \includegraphics[width=16cm]{../figures/main_03b.png}
    \centering
    \caption{False positive rates for spurious correlations \emph{increase} with sample size. \textit{Left:} Spearman rank correlations and 95\% bootstrap confidence intervals between learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. The thick dashed lines indicate the threshold for statistical significance for the Spearman correlation at the corresponding sample size. Markers are jittered along the x-axis for legibility. \textit{Right:} False positive rates for learning rate asymmetry ($\kappa$) and depression scores (7-down) as a function of sample size and proportion of C/IE participants. False positive rate was calculated as the proportion of bootstrap samples in which the Spearman rank correlation between $\kappa$ and 7-down was statistically significant ($p<0.05$, two-sided). The horizontal dotted line denotes the expected false positive rate at $\alpha = 0.05$.}
    \label{fig:fig04}
\end{figure}

### 2023-04-27 manuscripts/supp.tex
reupload

Participants were recruited as part of two independent studies to investigate reward processing in individuals with and without a history of major depression and bipolar disorder (results of those studies not reported here). The studies were approved by the Institutional Review Board of Rutgers University (\#2019000738, \#2018000629), and all participants provided informed consent. Participants received monetary compensation for their time (rate USD \$12/hr), plus an incentive-compatible bonus based on task performance.

@@

Here we sought to examine whether actual psychiatric patients were equally or more likely to fail infrequency items than healthy controls in settings similar to those experienced by online participants (i.e., completing an experiment online, on a computer in one's home or otherwise chosen environment). Although the small sample precludes any definitive conclusion, it is noteworthy that the model least consistent with the data was the one where psychiatric patients were more likely to fail infrequency-item attention checks. Indeed, a model in which healthy controls failed attention checks at a greater rate than patients was credibly preferred to the alternative. This preliminary finding may reflect differences in motivation between patients and controls for participating in psychiatric research. Indeed, whereas healthy controls may be primarily motivated to participate for monetary purposes, patients may be motivated to participate to further scientific research that may ultimately benefit them (or others suffering from the same conditions). That is, patients may have more ``stakes in the game,'' and therefore may be motivated to provide higher-quality responses. Regardless, further research is required to examine whether this preliminary finding holds in larger samples and other testing contexts.

>>>>

Participants were recruited as part of two independent studies to investigate reward processing in individuals with and without a history of major depression and bipolar disorder (results of those studies not reported here). The studies were approved by the Institutional Review Board of Rutgers University (\#2019000738, \#2018000629), and all participants provided informed consent. Participants received monetary compensation for their time (rate USD \$20/hr), plus an incentive-compatible bonus based on task performance.

@@

Here we sought to examine whether actual psychiatric patients were equally or more likely to fail infrequency items than healthy controls in settings similar to those experienced by online participants (i.e., completing an experiment online, on a computer in one's home or otherwise chosen environment). Although the small sample precludes any definitive conclusion, it is noteworthy that the model least consistent with the data was the one where psychiatric patients were more likely to fail infrequency-item attention checks. Indeed, a model in which healthy controls failed attention checks at a greater rate than patients was credibly preferred to the alternative. This preliminary finding may reflect differences in motivation between patients and controls for participating in psychiatric research. Indeed, whereas healthy controls may be primarily motivated to participate for monetary purposes, patients may be motivated to participate to further scientific research that may ultimately benefit them (or others suffering from the same conditions). That is, patients may have more ``stakes in the game,'' and may therefore be more motivated to provide higher-quality responses. Regardless, further research is required to examine whether this preliminary finding holds in larger samples and other testing contexts.
